r---
title: 'CMTH 642 Data Analytics: Advanced Methods'
output:
  word_document: default
  pdf_document: default
  html_document: default
---
<center> <h1> Assignment 1 (10%) </h1> </center>
<center>  <h3> Ramello Peralta </h2> </center>
<center> <h3> CMTH642 D10 500519802 </h2> </center>
---

#### 1. Read the csv files in the folder. (3 points)
```{r}
micronutrients.data <- read.csv("C:/Users/Ramello/Documents/CMTH642/Assignment1/USDA_Micronutrients.csv", header = T, sep = ",", na.strings = c("", "!", "NA"))

macronutrients.data <- read.csv("C:/Users/Ramello/Documents/CMTH642/Assignment1/USDA_Macronutrients.csv", header = T, sep = ",", na.strings = c("", "!", "NA"))

```

#### 2. Merge the data frames using the variable "ID". Name the Merged Data Frame "USDA". (6 points)
```{r}
nutrition.data <- merge(micronutrients.data, macronutrients.data, by="ID")
```

#### 3. Check the datatypes of the attributes. Delete the commas in the Sodium and Potasium records. Assign Sodium and Potasium as numeric data types. (6 points)
```{r}
str(nutrition.data)#sodium and potassium are char types

nutrition.data$Sodium <- as.numeric(gsub(",","", nutrition.data$Sodium)) #gsub outputs char, which we convert to numeric with as.numeric.
nutrition.data$Potassium <- as.numeric(gsub(",","", nutrition.data$Potassium))

str(nutrition.data$Sodium); str(nutrition.data$Potassium);
```

#### 4. Remove records (rows) with missing values in more than 4 attributes (columns). How many records remain in the data frame? (6 points)
```{r}
nutrition.data$na4 <- apply(is.na(nutrition.data), 1, sum) #na4 column counts number of NA per row
sum(nutrition.data$na4 > 4) #170 records

nu.data <- subset(nutrition.data, nutrition.data$na4 <= 4)
str(nu.data)

#6887 records remain. 170 records had more than 4 NA values.
```

#### 5. For records with missing values for Sugar, Vitamin E and Vitamin D, replace missing values with mean value for the respective variable. (6 points)
```{r}
sum(is.na(nu.data$Sugar)); sum(is.na(nu.data$VitaminE)); sum(is.na(nu.data$VitaminD)); #NA before

for(c in c(4,9,10)){
  nu.data[,c][is.na(nu.data[,c])] <- (mean(nu.data[,c], na.rm = T))
}

sum(is.na(nu.data$Sugar)); sum(is.na(nu.data$VitaminE)); sum(is.na(nu.data$VitaminD)); #checking if NAs were replaced

```

#### 6. With a single line of code, remove all remaining records with missing values. Name the new Data Frame "USDAclean". How many records remain in the data frame? (6 points)
```{r}
USDAclean <- na.omit(nu.data)

str(USDAclean)
#6310 records remain
```

#### 7. Which food has the highest sodium level? (6 points)
```{r}
USDAclean[which.max(USDAclean$Sodium),]
USDAclean$Description[which.max(USDAclean$Sodium)]
#Table salt has the highest sodium level

```

#### 8. Create a histogram of Vitamin C distribution in foods. (6 points)
```{r}
hist(USDAclean$VitaminC, breaks = 20)
#many outliers
hist(USDAclean$VitaminC[which(USDAclean$VitaminC < 100)], breaks=20, main = "Histogram VitaminC < 100", xlab = "VitaminC") #omitting high > 100 VitaminC foods for better visibility on lower end of distribution
```

#### 9. Create a boxplot to illustrate the distribution of values for TotalFat, Protein and Carbohydrate. (6 points)
```{r}
str(USDAclean[,c(13:15)])
summary(USDAclean[,c(13:15)])
boxplot(USDAclean[,c(13:15)], col=c("brown","orange","darkolivegreen"),  main="Distribution of Protein, TotalFat and Carbs")

boxplot(USDAclean[,c(13:15)], col=c("brown","orange","darkolivegreen"),  main="Distribution of Protein, TotalFat and Carbs w/o outliers", outline=FALSE) 

```

#### 10. Create a scatterplot to illustrate the relationship between a food's TotalFat content and its Calorie content. (6 points)
```{r}
scatter.smooth(USDAclean$TotalFat~USDAclean$Calories, xlab = "Calories", ylab = "TotalFat", main = "TotalFat vs Calorie content")
#Fat and Calories look to be positively correlated

```

#### 11. Add a variable to the data frame that takes value 1 if the food has higher sodium than average, 0 otherwise. Call this variable HighSodium. Do the same for High Calories, High Protein, High Sugar, and High Fat. How many foods have both high sodium and high fat? (8 points)
```{r}
USDAclean$HighSodium <- 0
USDAclean$HighCalories <- 0
USDAclean$HighProtein <- 0
USDAclean$HighSugar <- 0
USDAclean$HighFat <- 0

k <- 17 #[,17] is the index for HighSodium column, k increases by 1 on each loop to change High column
for (c in c(2,12,13,4,14)) { #each column corresponds to a High column at the end hence they are sequenced in this way
  for (i in 1:length(USDAclean[,1])) {
        if(USDAclean[,c][i] > mean(USDAclean[,c], na.rm=T)){
            USDAclean[,k][i] <- 1
        } 
  } 
  k = k+1
}

# a <- 17# check that the output into each High columns are correct
# for (c in c(2,12,13,4,14)) {
#   Check = 0
#   for (i in 1:length(USDAclean[,1])){
#     if (((USDAclean[,c][i] < mean(USDAclean[,c])) & (USDAclean[,a][i] == 1)) |
#         ((USDAclean[,c][i] > mean(USDAclean[,c])) & (USDAclean[,a][i] == 0))) {
#         Check = Check + 1
#     }
#   }
#   a = a+1
#   print(Check) 
# }#Check should be 0 on each loop


nrow(subset(USDAclean, (USDAclean$HighSodium == 1 & USDAclean$HighFat == 1)))
#644 food items have both high sodium and high fat
```

#### 12. Calculate the average amount of iron, for high and low protein foods. (8 points)
```{r}
aggregate(USDAclean$Iron~USDAclean$HighProtein, USDAclean, mean)
mean(subset(USDAclean, (USDAclean$HighProtein == 1))$Iron)
mean(subset(USDAclean, (USDAclean$HighProtein == 0))$Iron)
#Avg iron in high protein = 3.06
#Avg iron in low protein = 2.69
```

#### 13. Create a script for a "HealthCheck" program to detect unhealthy foods. Use the algorithm flowchart below as a basis for this script. (8 points)
```{r}
library(jpeg)
require(jpeg)
img<-readJPEG("HealthCheck.jpg")
plot(1:4, ty = 'n', ann = F, xaxt = 'n', yaxt = 'n')
rasterImage(img,1,1,4,4)
```
```{r}
# pseudocode based on flow chart:
# if food has high sodium, then check sugar, else pass.
# if food has high sugar, then check fat, else pass.
# if food as high fat, then fail, else pass

# according to the flow chart this is a binary classification and a food is classified as unhealthy
# if it contains high fat, sugar and sodium. We already have columns for high sodium, sugar and fat.
# the problem can be solved with the expression (HighSodium == 1 & HighFat == 1 & HighSugar == 1)

# for (i in 1:length(USDAclean[,1])){
#   if(HighSodium[i] == 1 & HighFat[i] == 1 & HighSugar[i] == 1){
#   Fail[i] <- 1
#   }
# }
```

#### 14. Add a new variable called HealthCheck to the data frame using the output of the function. (8 points)
```{r}
USDAclean$HealthCheck <- 0

for (i in 1:length(USDAclean[,1])) {
  if(USDAclean$HighSodium[i] == 1 & USDAclean$HighFat[i] == 1 & USDAclean$HighSugar[i] == 1){
    USDAclean$HealthCheck[i] = 1 
  }
}

USDAclean$HealthCheck <- factor(USDAclean$HealthCheck, labels = c("PASS", "FAIL"))
```

#### 15. How many foods in the USDAclean data frame fail the HealthCheck? (8 points)
```{r}
head(subset(USDAclean, (USDAclean$HealthCheck == "FAIL")))
nrow(subset(USDAclean, (USDAclean$HealthCheck == "FAIL")))
#237 foods failed the health check
```

#### 16. Save your final data frame as "USDAclean_ [your last name]." (3 points)
```{r}
USDAclean_Peralta <- USDAclean
```


This is the end of Assignment 1

Ceni Babaoglu, PhD
