---
title: 'CMTH 642 Data Analytics: Advanced Methods'
output:
  html_document: default
  pdf_document: default
  word_document: default
---
<center> <h1> Assignment 3 (10%) </h1> </center>
<center>  <h3> Ramello Peralta </h2> </center>
<center> <h3> CMTH642 D10 500519802 </h2> </center>
---


#### 1. Import to R the following fiel: http://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv (The dataset is related to white Portuguese "Vinho Verde" wine. For more info: https://archive.ics.uci.edu/ml/datasets/Wine+Quality) (3 points)
```{r}
wine <- read.csv(file="http://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv", sep=";")

#install.packages("class")
#install.packages("gmodels")
#install.packages("caret")
#library(caret)
library(class)
library(gmodels)
```

#### 2. Check the datatypes of the attributes. (3 points)
```{r}
str(wine)
summary(wine)

```
  
#### 3. Are there any missing values in the dataset? (4 points)
```{r}
sum(is.na(wine))
#0 NA values
```

#### 4. What is the correlation between the attributes other than Quality? (10 points)
```{r}
ncol(wine) #quality is the 12th attribute
cor(wine[,-12])
```

#### 5. Graph the frequency distribution of wine quality by using Quality. (10 points)
```{r}
hist(wine$quality, xlab = "Quality Score", main = "Distribution of Wine Quality")
```

#### 6.	Reduce the levels of rating for quality to three levels as high, medium and low. Assign the levels of 3 and 4 to level 0; 5 and 6 to level 1; and 7,8 and 9 to level 2. (10 points)
```{r}
wine$quality<- factor(wine$quality, ordered = T)
wine$rating <- ifelse(wine$quality < 5, 0, ifelse(wine$quality < 7, 1, 2))
wine$rating <- factor(wine$rating, levels = c(0, 1, 2), labels = c("low", "medium", "high"))

table(wine$rating)

wine <- wine[-12]#we don't need the quality column anymore because it was replaced with rating factor
head(wine)
```

#### 7. Normalize the data set by using the following function: (12 points)
```{r}
normalize <- function(x){
  return ((x - min(x)) / (max(x) - min(x)))
}
```

```{r}
winenorm <- as.data.frame(lapply(wine[1:11], normalize)) #rating = col 12
winenorm <- cbind(wine$rating, winenorm) #rebinded the labels after normalization to col1
summary(winenorm)#all numerical variables' min/max are 0/1 respectively meaning the dataset has been normalized
```

#### 8. Divide the dataset to training and test sets. (12 points)
```{r}
set.seed(1)
index <- sample(1:nrow(winenorm), 0.75*nrow(winenorm))
wine_train <- winenorm[index,]
wine_test <- winenorm[-index,]
```

#### 9. Use the KNN algorithm to predict the quality of wine using its attributes. (12 points)
```{r}
wine_train_labels <- wine_train[,1]
wine_test_labels <- wine_test[,1]

wine_test_pred <- knn(train = wine_train[,2:12],
                      test = wine_test[,2:12],
                      cl = wine_train[,1], k = 70) #k = sqrt of total records
```

#### 10. Display the confusion matrix to evaluate the model performance. (12 points)
```{r}
cmatrix <- (table(Predicted = wine_test_pred, Actual = wine_test_labels))
cmatrix
```

#### 11. Evaluate the model performance by computing Accuracy, Sensitivity and Specificity. (12 points)
```{r}
CrossTable(wine_test_labels, wine_test_pred, prop.chisq = FALSE)
#sens = TP/(TP + FN)
#spec = TN/(TN + FP)
specLow = (870+192+50+81)/(870+192+50+81+0)
sensLow = 0
specMedium = (81+1)/(81+1+31+192)
sensMedium = 870/(870+50)
specHigh = (31+870)/(31+870+1+50)
sensHigh = 81/(81+192)

sensitivity <- c(sensLow, sensMedium, sensHigh)
specificity <- c(specLow, specMedium, specHigh)

sm <- data.frame(sensitivity)
row.names(sm) <- c("low", "medium", "high")
cbind(specificity, sm)

totalacc = (870+81) / (870+81+31+1+50+192); totalacc

```

This is the end of Assignment 3

Ceni Babaoglu, PhD